---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

## Setup
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(lubridate)
library(lattice)
```

## Loading and preprocessing the data
```{r import}
setwd("~/Documents/Classes/DataSciCourse/RepData_PeerAssessment1")
data <- read.csv(unz('activity.zip', "activity.csv"), colClasses = c("numeric", 
                                                "character", "character"))
data <- as_tibble(data)
data
```

Date was imported as a character. Let's fix that and the intervals. The intervals 
are in the following base 60 format: HHMM where H is an hour digit and M is a 
minute digit. This makes a linear plot of the interval data skip large sections 
of base 10 space. Here are the date conversions:
```{r date}
# This 'for' loop converts the intervals to raw minutes
data <- data %>% mutate(Minutes = NA)
for(i in 1:dim(data)[1]) {
        if(nchar(data[i, 3]) <= 2){
                data$Minutes[i] = as.numeric(data[i, 3])
        } else if(nchar(data[i, 3]) == 3){
                data$Minutes[i] = 60*as.numeric(substr(data[i, 3], 1, 1)) + 
                                   as.numeric(substr(data[i, 3], 2, 3))
        } else if(nchar(data[i, 3]) == 4){
                data$Minutes[i] = 60*as.numeric(substr(data[i, 3], 1, 2)) + 
                                   as.numeric(substr(data[i, 3], 3, 4))
        }
}

# This 'for' loop will reformat the intervals to be "HH:MM". Some of the 
#       intervals have only one, two, or three digits and this must be taken
#       into account with the 'if' statements.
for(i in 1:dim(data)[1]) {
        if(nchar(data[i, 3]) == 1){
                data[i, 3] = paste("00:0", data[i, 3], sep = "")
        } else if(nchar(data[i, 3]) == 2){
                data[i, 3] = paste("00:", data[i, 3], sep = "")
        } else if(nchar(data[i, 3]) == 3){
                data[i, 3] = paste("0", substr(data[i, 3], 1, 1), ":", 
                                   substr(data[i, 3], 2, 3), sep = "")
        } else if(nchar(data[i, 3]) == 4){
                data[i, 3] = paste(substr(data[i, 3], 1, 2), ":", 
                                   substr(data[i, 3], 3, 4), sep = "")
        }
}

data <- data %>% mutate(date_time = as.POSIXct(paste(date, interval), 
                       format = "%Y-%m-%d %H:%M"), date = ymd(date))

data
```

## What is mean total number of steps taken per day?
Let's group by day and take some summary statistics.
```{r DayGrouping}
data <- group_by(data, date)
sum_steps <- summarise(data, sum = sum(steps, na.rm = TRUE))

# Storing the mean and median for future calls.
avg <- mean(sum_steps$sum)
med <- median(sum_steps$sum)

# The histogram. 30 breaks looks better than the default.
hist(sum_steps$sum, breaks = 30, main = 'Histogram of Total Steps Taken in 61 Days', xlab = 'Total Steps Taken')
box()
```

The mean number of steps taken each day is `r signif(avg, 4)` and the median is 
`r as.integer(med)`.

## What is the average daily activity pattern?
```{r}
data <- group_by(data, Minutes)
avg_steps <- summarize(data, mean =  mean(steps, na.rm = TRUE))
maximum <- max(avg_steps$mean)
max_interval <- avg_steps[avg_steps$mean == maximum, 1]

plot(avg_steps$Minutes/60, avg_steps$mean, type = 'l', lwd = 2, 
     xlab = 'Hours of the day', ylab = 'Average number of steps', 
     main = 'Average steps across 61 days\nand measured in 5 minute intervals')
```

The maximum number of steps in any five minute interval was 
`r signif(maximum, 3)` steps which started at 
`r paste(max_interval %/% 60, 'hours', max_interval %% 60, 'minutes')`.

## Imputing missing values
```{r Imputation}
# Filter for rows with NA steps and get the number of rows.
missing <- dim(
        filter(data, is.na(steps))
)[1]

# Replace missing values with the average for that interval
filled_data <- data
for(i in 1:dim(data)[1]){
        if(is.na(data[i, 1])){
                filled_data[i,1] <- 
                avg_steps[avg_steps$Minutes==data$Minutes[i], 'mean']
        }
}


filled_sum_steps <- filled_data %>% 
        group_by(date) %>%
        summarise(sum = sum(steps, na.rm = TRUE))

par(mfcol = c(1,2))
hist(sum_steps$sum, breaks = 30, main = 'Raw Histogram of Total\nSteps Taken in 61 Days', xlab = 'Total Steps Taken\n(Raw Data)')
box()
hist(filled_sum_steps$sum, breaks = 30, main = 'Filled Histogram of Total\nSteps Taken in 61 Days', xlab = 'Total Steps Taken\n(Filled Data)')
box()
```

There are `r missing` missing values in the 'steps' column. They were replaced 
with the mean value for the given interval. Eight days that previously 
had a total of 0 steps get pushed closer to a total of 10,000 steps. This causes 
the mean frequency to become inflated.

## Are there differences in activity patterns between weekdays and weekends?
```{r Weekdays}
# First we need to group by Weekday and by Minutes, then take the avg of steps.
filled_data$Weekdays <- factor(x = rep(NA, dim(filled_data)[1]), 
                               levels = c("Weekday", "Weekend"))
MTWRF <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
for(i in 1:dim(filled_data)[1]){
        if(weekdays(filled_data$date[i]) %in% MTWRF){
                filled_data$Weekdays[i] <- as.factor("Weekday")
        } else {
                filled_data$Weekdays[i] <- as.factor("Weekend")
        }
}

filled_avg_steps <- filled_data %>% 
        group_by(Weekdays, Minutes) %>%
        summarise(Mean = mean(steps))

xyplot(Mean ~ Minutes | Weekdays, data = filled_avg_steps, type = 'l', 
       xlab = "Interval in minutes", ylab = "Mean number of steps", 
       main = "Average steps in a given interval, by weekday/end", 
       layout = c(1,2))
```

Weekdays have a larger spike of activity close to 9 am.